---
title: "Introduction to PartMI"
author: "Dany Mukesha"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to PartMI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

## What is Part Mutual Information?

Part Mutual Information (PMI) quantifies the direct association between two variables $X$ and $Y$ after removing the influence of a set of conditioning variables $Z$:

$$
PMI(X; Y | Z) = \sum_{x,y,z} P(x,y,z) \log \frac{P(x,y|z)}{P(x|z)P(y|z)}
$$

This is mathematically equivalent to Conditional Mutual Information (CMI), but applied specifically to the problem of direct association network inference.

### Why not just use correlation?

Consider three genes A, B, C where A regulates B and B regulates C. Standard correlation or mutual information will show a false positive association between A and C (indirect edge). PMI conditioning on B correctly identifies no direct A-C link.

## Installation

```{r, eval = FALSE}
remotes::install_github("danymukesha/PartMI")
```

## Basic Usage

### Mutual Information

```{r mi_example}
library(PartMI)
set.seed(42)

# Independent variables
x <- rnorm(200)
y <- rnorm(200)
mi(x, y)  # near zero

# Dependent variables
y_dep <- x + rnorm(200, 0, 0.3)
mi(x, y_dep)  # positive value
```

### Part Mutual Information

```{r pmi_example}
# Common cause scenario
z <- rnorm(200)
x <- 0.7 * z + rnorm(200, 0, 0.5)
y <- 0.7 * z + rnorm(200, 0, 0.5)

# MI shows spurious association
mi_xy <- mi(x, y)
mi_xy

# PMI removes the common cause
pmi_xy <- pmi(x, y, z)
pmi_xy
```

### Network Inference

```{r network_example}
# Simulate a small regulatory network
n_obs <- 100
TF1  <- rnorm(n_obs)
TF2  <- rnorm(n_obs)
Gene1 <- 0.8 * TF1 + rnorm(n_obs, 0, 0.4)
Gene2 <- 0.6 * TF2 + rnorm(n_obs, 0, 0.4)
Gene3 <- 0.5 * Gene1 + 0.3 * TF2 + rnorm(n_obs, 0, 0.3)
Noise <- rnorm(n_obs)

dat <- data.frame(TF1, TF2, Gene1, Gene2, Gene3, Noise)

net <- pmi_network(dat, n_permutations = 50, threshold = 0.05)

# View the adjacency matrix
print(net$adjacency)

# Summary
print(net)
```

## Estimation Methods

### Binning (default)

Fast histogram-based estimator. Good for exploratory analysis. Control granularity with the `bins` parameter.

```{r binning}
pmi(x, y, z, method = "binning", bins = 10)
```

### k-Nearest Neighbor (kNN)

More accurate, especially for higher-dimensional Z. Based on the Frenzel-Pompe estimator. Control bias-variance tradeoff with `k`.

```{r knn}
pmi(x, y, z, method = "knn", k = 5)
```

## Preprocessing

```{r preprocessing}
normalize_data(dat, method = "zscore")
normalize_data(dat, method = "rank")
normalize_data(dat, method = "quantile")
discretize_data(dat, bins = 5)
```

## References

1. Zhao et al. (2016) "Part mutual information for quantifying direct associations in networks." *PNAS* 113(18): 5130-5135.
2. Frenzel & Pompe (2007) "Partial Mutual Information for Coupling Analysis of Multivariate Time Series." *Physical Review Letters* 99(20): 204101.
3. Kraskov, Stögbauer & Grassberger (2004) "Estimating mutual information." *Physical Review E* 69(6): 066138.
